---
title: Main Analyses
format:
  html:
    code-fold: true
    code-summary: "code"
    embed-resources: true
    sidebar:
      collapse-level: 3
      pinned: true
toc: true
params:
  frailty: "off"
bibliography: references.bib
---

```{r}
#| label: setup
#| include: false
#| file: mort_functions.R
library(Statamarkdown)
library(ggplot2)
```

```{r}
#| echo: false
hide_frailty <- "#"
cat("Re-running frailty models currently set to", params$frailty)
myplot<-function(dag_name) {
  library(ggplot2)
  ggplot(dag_name,aes(
      x = x,
      y = y,
      xend = xend,
      yend = yend
    )) +
    geom_dag_point(text=FALSE,color="darkgrey") +
    geom_dag_label_repel(aes(label = label, fill = label), 
      col = "black", show.legend = FALSE, max.overlaps=Inf) +
    geom_dag_edges(aes(edge_alpha=0.5),edge_color = "black") +
    theme_dag()
}
```

Last run October 25, 2024 at 2:54PM-3:00PM (check modified date saved results)

::: {.callout-note collapse="true"}
## Functions used in the analytic code below

```{r}
#| echo: false
#| warning: false
#| class-output: "sourceCode r"

cat(readLines("mort_functions.R"), sep = "\n")
```

(The algorithms for these functions used in the analysis and some of the manipulation of the results were based on examples from [@balan2020]. Any errors identified in adapting them or in their application to our example are ours alone.)
:::

::: {.callout-note collapse="true"}
## Dataset

Stata code used to create the analytic data set of all 5 cancers from file of SEER variables distributed by project data steward

```{stata}
#| eval: false
use "data/mort imputed 1013.dta"
mi extract 0
* reorder to match Paul's outputs
recode cancer_sitegroup  (3=1 "breast, triple negative") ///
  (2=2 "breast, HER2+") (1=3 "breast, ER/PR+ HER2-") (4=4 "colorectal") ///
  (5=5 pancreas),gen(cancer_site1)
* surgcat has 5 categories should have y/n by Paul's outputs
gen surgcat=surg>0
recode sexrecode (1 9 =1 male) (2=0 female),gen(sex) label(male_sex)
label values sex male_sex
label var sex "Sex"
tostring yeardx,gen(year_str)
encode year_str,gen(yeardx_cat)
label define yn 1 "Yes" 0 "No"
label values married surgcat chemo radiation her2therapy yn
recode colon_rectum (1=0) (2=.) (3=1),gen(colon_rectal)
label define colon_rectal 1 rectal 0 colon 
label values colon_rectal colon_rectal
label var colon_rectal "Colorectal type"
compress
qui dtable i.(pvgroup race age_group poverty rurality married yeardx_cat stage ///
                surgcat chemo radiation state grade her2therapy colon_rectum) ///
              ,by(cancer_site1,nototals) export(table1.html,replace)
* switch to study window end 12/31/2021
stset, clear
stset time22, failure(death22)
gen survtime=_t
gen event=_d
save "data/mort",replace
de pvgroup
```
:::

## Overview of modeling strategy {#model-strat}

We used directed acyclic graphs to justify the selection of variables for the analysis.[@pearl1995; @lipsky2022]

The primary predictor of interest is the gene group describing the presence or absence of specific pathogenic variants. The outcome is mortality. In addition the following variable groups were included:

__Covariate list__

| Clinical | Treatment | Shared ancestry | Access and quality | Selection into cohort |
|---------------|---------------|---------------|---------------|---------------|
| Age at diagnosis | Chemotherapy | Race | Poverty ( county level) | Genetic testing |
| Stage | Surgery |  | Rurality (county level) |  |
| Grade (cancer-specific) | Radiation |  | State  |  |
| Rectal vs colon (cancer-specific) | Receipt of HER2-therapy (cancer-specific) |  |  |  |

As only people from the SEER registry who had genetic testing within 6 months of diagnosis were included, all of the results are conditioned on receipt of genetic testing even though this variable is not formally included in our analysis.

::: {.callout-note collapse="true"}
### Variable selection - rationale (click to view) 

The goal of the analysis is to describe whether people who have one of several pathogenic variants (PVs) or gene groups would be predicted to have a different mortality hazard in comparison to the average hazard of people with the cancer. More specifically, for a person who has been diagnosed and is planning or undergoing treatment with a cancer of known stage and grade and then finds out that they have a PV, the pertinent question becomes whether this finding will have any implications for their mortality risk beyond what they already know about their diagnosis.

We could expect that the presence of a PV can potentially modify the age at which one gets cancer, the stage or grade at which it occurs and the treatment that is received. These clinical and treatment variables are then mediators between the presence of a PV and mortality as represented in the DAG below. However, given that a patient already knows that information, to answer the question described requires trying to estimate any direct effect of PV on mortality independent of any paths through the described mediators and any possible confounders of the relationship.

```{r}
#| warning: false
library(dagitty)
library(ggdag)
risk<-dagify(
  Y ~ PV + C + M + C2, PV ~ C, M ~ PV + C2, 
  coords =list(
    x=c(Y=7,C=3, PV=0, M =3 ,C2 = 7 ), 
    y=c(Y=0,C=-3,PV=0, M = 3, C2 = 3  )
    ),
  exposure="PV",
  outcome="Y",
  labels=c(PV="PV", C= "PV-outcome \n Confounder" , Y=" Mortality", M = "Clinical/treatment Mediator",
           C2="Mediator-outcome \n confounder")
) 
myplot(risk)
```

We wish to block all the clinical/treatment mediators as well as any confounders of the PV - mortality relationship in order to test for whether there is any direct effect of PV on mortality. Unfortunately, once we adjust for the mediators we need to also account for any mediator - outcome and any PV - mediator confounders as well. If we do not then the correlations in that pathway will be included in our estimate of the PV - mortality relationship.[@elwert2014]

In the more detailed DAG below we show the mediators whose paths we wish to block (**age at diagnosis**, **stage/grade** and the treatment received which includes **chemotherapy**, **surgery** and **radiation**). We also include the receipt of ***HER2*****-therapy** as a disease specific treatment variable for breast cancer and ***Rectal vs Colon*** location for colorectal cancer.


```{r}
#| fig-height: 6
#| fig-width: 8
#| warning: false
risk<-dagify(
  Y ~  PV + M1 + M2 + M3 + C2 + R + YD, PV ~ R,
    M1 ~PV + C2, M2 ~ PV + C2, M3 ~ PV + C2,
    M3 ~ M1 + M2 + YD, R ~ C, 
    GT ~ YD + M1 + M2 + C2 + R,  
  coords =list(
    x=c(Y=7,C=1.5, R = 1.5, PV=0, M1 = 1.5 , M2 = 2.5 ,M3 = 3,C2 = 7,YD= 5 ,GT = 7), 
    y=c(Y=0,C=-3.5,R = -1.5, PV=0, M1 = 1, M2=3 , M3= 6 , C2 = 3, YD=6, GT = 5  )
    ),
  exposure="PV",
  outcome="Y",
  labels=c(PV="PV", C= "Shared Ancestry" , Y=" Mortality", M1 = "Age at diagnosis",
           M2 = "Stage/grade", M3="Treatment", C2="Mediator-outcome \n confounders",
           R = "race/ethnicity", YD = "year of diagnosis", GT = "gene testing")
) 

set.seed(7)
set.seed(9670)
myplot(risk)
```

There are not many confounders of the direct PV - mortality or PV - mediator relationships as in order to be a confounder it has to cause (requiring it to come before) the PV as well as the outcome. However, the most commonly raised potential confounder of gene outcome relationships is shared ancestry. The primary measure of shared ancestry that we have is the SEER **race/ethnicity** variable. Of the gene groups we have *BRCA1/2* is the only one with an obvious shared ancestry group of Ashkenazi Jews who share a founder mutation. However, apart from this one group it seems that *BRCA1/2* prevalence is not uncommon and quite similar across race and ethnic groups[@hall2009] and the latter seems true for the other gene groups. But we decided to still include this race variable as as a proxy for shared ancestry.

Again, the biggest problem with estimating the direct effect of PV on mortality is that once we have included the mediators we have to now block all the potential mediator-outcome confounders labeled collectively in the green box in the DAG. For example, given that we hypothesize that PVs may cause treatment changes, then any confounder of the treatment -\> mortality relationship, such as access or quality of care poverty which limits access to treatment, opens up a collider path with respect to the PV mortality relationship. Inclusion of treatment as a collider, while it blocks the mediating path from PV -\> Treatment -\> Mortality, it will open a path from PV -\> treatment \<- Poverty -\> Mortality. Thus we include the county level variables **poverty**, **rurality** as well as **state** as a proxy for quality and access related potential mediator-outcome confounders.

A second potential mediator-outcome confounder is the **year of diagnosis** which can cause treatment changes if cancer therapy changes and also might have an effect on outcome due to other changes in supportive care occur over time. As such it also has to be blocked to avoid a backdoor path from the treatment mediator to the outcome which would bias the PV - mortality relationship. An interesting feature of year of diagnosis is that selection into our cohort changed quite a bit from 2013-2019 as genetic testing rates increased significantly.[@kurian2023]  Thus year of diagnosis is causally related to genetic testing. Given the additional causal pathways from the clinical characteristics of the cancer and race to genetic testing, a pathway opens up from year of diagnosis to mortality as the implied genetic testing covariate is a collider on this pathway.[@elwert2014]  This explains an odd finding seen in the regression results which is a large decrease in the hazard ratio for later years of diagnosis as genetic testing spreads from the highest risk groups to lower risk groups and the cohorts include an increasing number of people with lower clinical risk.  

People do not get into the cohort unless they have a genetic test. This automatically includes **genetic testing** as a covariate whether we want it or not. Genetic testing is more common in higher risk cancers (early age of diagnosis or high stage and grade) and also potentially causally related to race/ethnicity as well as some of the access and quality variables and the relationships are changing over time. The concern of the selection mechanism would be if it introduced any backdoor paths between PV and mortality. However the DAG is getting a bit complicated to for us to easily detect this at this point.

Fortunately, once we have encoded the causal relationships we propose exist we can use functions that apply an axiomatic system for replacing probability formulas containing the causal relationships with ordinary conditional probabilities. That allows us to identify whether the adjustment set is able to evaluate a specific relationship without bias. This axiomatic system has the name ‘do-calculus’ invented by Judea Pearl.[@pearl2010] Below we can show that if we condition on the variables in the DAG above the relationship between PV and mortality are d-separated which would allow us to estimate the presence or absence of a direct effect.

```{r}
#| fig-height: 7
#| fig-width: 8
#| warning: false
risk1<-dagify(
  Y ~   M1 + M2 + M3 + C2 + R + YD, PV ~ R,
    M1 ~PV + C2, M2 ~ PV + C2, M3 ~ PV + C2,
    M3 ~ M1 + M2 + YD, R ~ C, 
    GT ~ YD + M1 + M2 + C2 + R,  
  coords =list(
    x=c(Y=7,C=1.5, R = 1.5, PV=0, M1 = 1.5 , M2 = 2.5 ,M3 = 3,C2 = 7,YD= 5 ,GT = 7), 
    y=c(Y=0,C=-3.5,R = -1.5, PV=0, M1 = 1, M2=3 , M3= 6 , C2 = 3, YD=6, GT = 5  )
    ),
  exposure="PV",
  outcome="Y",
  labels=c(PV="PV", C= "Shared Ancestry" , Y=" Mortality", M1 = "Age at diagnosis",
           M2 = "Stage/grade", M3="Treatment", C2="Mediator-outcome \n confounders",
           R = "race/ethnicity", YD = "year of diagnosis", GT = "gene testing")
) 
set.seed(3920)
ggdag::ggdag_dseparated(
  risk1,
  from = "PV",
  to = "Y",
  controlling_for = c("M1", "M2", "M3","R","C2","GT","YD"),
  text = FALSE,
  use_labels  = "label") +
  theme_dag() +
  labs(title = "If we add the confounders of all the mediators then PV and Y \n are independent as they should be if there is no direct path")
```

Based on this we can argue that we may be able to identify the direct effect of PV on mortality. However, in case it is not obvious the two biggest assumptions in the selection of covariates with respect to identifying the PV - mortality relationship are as follows:

1.  There are no unmeasured confounders beyond race that capture important shared ancestry features that would cause both the PV and higher mortality.
2.  The covariate set of county level variables poverty, rurality as well as state are an adequate proxy for the quality and access related potential mediator-outcome confounders.

Although not always clear, both of these assumptions are required for all gene outcome studies if they condition on any treatment or clinical severity variables, not just ours. However, we believe proposing a specific DAG provides a principled way of justifying the selection of covariates and allows us to start to try to identify the key variable sets for making these estimates. Perhaps most importantly it offers a transparent way for others to critique our choices and to propose and defend through additional analyses the need for other covariates.
:::

### Fixed effects model

The fixed effects modeled are estimated using the `coxph` procedure from the `Survival` package.[@therneau2024]

### Use of multilevel models to estimate gene group random effects

The implementation of multilevel models in survival data uses the term frailty models for a survival regression model that includes random effects. The term is often specifically used for those models that only include random effects for the intercept.[@austin2017] When the random effects are shared by a cluster the models are known as shared frailty models. Given the mathematical form of the survival model, the random intercept in the linear predictor has a multiplicative effect on the hazard function.

For our purposes we could propose a data-generating hypothesis for the use of multilevel models that involves viewing the pathogenic variants that have been found in our population as a random sample of all the possible genetic variations that can have an effect on the mortality of people who develop cancer. However, in this case we are primarily using a multilevel model as a strategy to avoid over-fitting. Given that we are fitting 8 gene groups over 5 cancers there are a total of 40 estimates of interest. There is heterogeneity in the size of the gene groups with some quite small. Multilevel modeling can be thought of "as adaptive regularization where the model itself tries to learn how skeptical it should be."[@mcelreath2020] In this sense it has some similarities to other strategies to avoid over-fitting such as ridge regression. The gamma distribution acts as a regularizing prior for the gene effects although in this case the parameters for that prior are learned from the data.

For estimation we used the profile EM (expectation-maximization) algorithm as implemented in the FrailtyEM package.[@balan2019]

## Breast Cancer Models

```{r}
#| code-fold: false
#| echo: false
covar1<-c("agedec","race","poverty","rurality","married","yeardx_cat","stage","surgcat","chemo",
          "radiation","state","her2therapy","grade","pvgroup")
```

### Triple negative

::: {.callout-note collapse="true"}
#### Sample description and missing data (click to view)

```{r}
#| warning: false
can_type=1
covar<-eval(as.name(paste0("covar",can_type,sep="")))
mort <-get_data(cancer_type=can_type) 
my_labels<-list(agedec ~ "Age (in decades from mean age)", race ~ "Race", 
                poverty ~ "Poverty", rurality ~ "Rurality", married ~ "Married", 
                yeardx_cat ~ "Year of diagnosis", stage ~ "Stage", 
                surgcat ~ "Surgery (Yes=1)", chemo ~ "Chemotherapy (Yes=1)", 
                radiation ~ "Radiation (Yes=1)", state~ "State", 
                her2therapy ~ "HER2 treatment (Yes=1)", grade ~ "Grade", 
                pvgroup ~ "Gene group")
gtsummary::tbl_summary(mort,label=my_labels, missing_text="missing",include=covar) |>
    gtsummary::modify_caption("Table 1. Covariates - Triple negative") |>
    gtsummary::modify_footnote(label = "Omitted category listed first for categorical variables" )
```
:::

#### Model output

##### Random effects for gene group (as shown in figure 1)

```{r}
#| message: false
#| warning: false
# random effects
if(params$frailty=="on") {
  m_emfrail<-get_random(mort,can_type,covar)
}

# fixed effects
m_fixed<-get_fixed(mort,covar)

# graph
## process fixed effects
fe<-process_fixed(m_fixed)
## process random effects
file_name=paste("data/","m_emfrail",can_type,".Rdata",sep="")
load(file_name)
logz<-process_random(m_emfrail)
## combine and graph
datt<-combine_effects(fe,logz)
p1<-graph_effects_vert(datt)
p1 <- forestploter::edit_plot(p1,
               col = 1,
               row = c(1:3,7),
               gp = gpar(fontface = "italic"))
p1 <- forestploter::insert_text(p1, text = "Triple-Negative Breast Cancer",
              part = "header",
              col = 1:3,
              gp = gpar(fontface = "bold"))
my_pval<-1/2*pchisq(frailtyEM:::summary.emfrail(m_emfrail)$loglik[[3]],1, lower.tail=FALSE)
my_chi<-paste0("Likelihood ratio test for heterogeneity \U{03A7}\U{00B2} = ",
               round(frailtyEM:::summary.emfrail(m_emfrail)$loglik[[3]],2),", p",pvalr(my_pval))
p1 <- forestploter::insert_text(p1, text = my_chi,
   part = "header", just = "right",
   col = 3, gp = gpar(fontsize=10) )

ggsave(p1,filename=paste("img/","m_emfrail",can_type,".","svg",sep=""))
ggsave(p1,filename=paste("img/","m_emfrail",can_type,".","pdf",sep=""),device=cairo_pdf)
```

![](img/m_emfrail1.svg)

::: {.callout-note collapse="true"}
##### Fixed effects model (click to view)

```{r}
gtsummary::tbl_regression(m_fixed,exponentiate=TRUE,label=my_labels)
```

Note the decreasing hazard ratio with year of diagnosis that is a product of the spread of genetic testing to lower clinical risk groups as discussed as part of the variable selection rationale in @model-strat above.
:::

##### Tests of heterogeneity

```{r}
fvar<-format(as.list(frailtyEM:::summary.emfrail(m_emfrail)$fr_var),scientific=FALSE)
ca<-format(as.list(frailtyEM:::summary.emfrail(m_emfrail)$ca_test),scientific=FALSE)
lrt<-format(as.list(frailtyEM:::summary.emfrail(m_emfrail)$loglik),scientific=FALSE)

p11<-table_fig(fvar,tbl_title="Variance",.9)
p12<-table_fig(ca,tbl_title="Commenges-Andersen test",1)
p13<-table_fig(lrt,tbl_title="Likelihood ratio test",.8)
cowplot::plot_grid(p11,p12,p13,nrow=1)
```

### Breast *HER2+*

::: {.callout-note collapse="true"}
#### Sample description and missing data (click to view)

```{r}
covar2<-covar1
can_type=2
covar<-eval(as.name(paste0("covar",can_type,sep="")))
mort <-get_data(cancer_type=can_type)
# mylabels as defined above for triple negative dz
gtsummary::tbl_summary(mort,label=my_labels, missing_text="missing",include=covar) |>
    gtsummary::modify_caption("Table 1. Covariates - HER2+") |>
    gtsummary::modify_footnote(label = "Omitted category listed first for categorical variables" )
```
:::

#### Model output

##### Random effects for gene group (as shown in figure 1)

```{r}
#| message: false

# random effects
if(params$frailty=="on") {
    m_emfrail<-get_random(mort,can_type,covar)  #random w. Lynch, fixed w.out
}

# fixed effects
levels(mort$pvgroup)<-c("Negative","ATM","BRCA 1/2","CHEK2","Other PV","Other PV","PALB2","VUS only") # Lynch(4) combined w other(6)
m_fixed<-get_fixed(mort,covar)

# graph
## process fixed effects
fe<-process_fixed(m_fixed)
## process random effects
file_name=paste("data/","m_emfrail",can_type,".Rdata",sep="")
load(file_name)
logz<-process_random(m_emfrail)
## combine and graph
datt<-combine_effects(fe,logz)  |>
      mutate(type= replace_na(type,"fixed effects"))
p2<-graph_effects_vert(datt)
p2 <- forestploter::edit_plot(p2,
               col = 1,
               row = c(1:3,7),
               gp = gpar(fontface = "italic"))
p2 <- forestploter::insert_text(p2, text = expression(bold(paste(bolditalic("HER2"),"-Positive Breast Cancer"))),
              part = "header",
              col = 1:3)
my_pval<-1/2*pchisq(frailtyEM:::summary.emfrail(m_emfrail)$loglik[[3]],1, lower.tail=FALSE)
my_chi<-paste0("Likelihood ratio test for heterogeneity \U{03A7}\U{00B2} = ",round(frailtyEM:::summary.emfrail(m_emfrail)$loglik[[3]],2),", p",pvalr(my_pval))
p2 <- forestploter::insert_text(p2, text = my_chi,
   part = "header", just = "right",
   col = 3, gp = gpar(fontsize=10) )

ggsave(p2,filename=paste("img/","m_emfrail",can_type,".","svg",sep=""))
ggsave(p2,filename=paste("img/","m_emfrail",can_type,".","pdf",sep=""),device=cairo_pdf)
```

![](img/m_emfrail2.svg)

The fixed effects cox model does not converge to stable estimates for the pvgroup when the Lynch genes are included. There are relatively few observations (25) in those with *HER2+* breast cancer. Thus Lynch was collapsed into the "Other PV" category for the fixed effects analysis. That is why the estimate is missing in the above figure.

::: {.callout-note collapse="true"}
##### Fixed effects model (click to view)

```{r}
my_labels<-list(agedec ~ "Age (in decades)", race ~ "Race", poverty ~ "Poverty",
                rurality ~ "Rurality", married ~ "Married", yeardx_cat ~ "Year of diagnosis",
                stage ~ "Stage", surgcat ~ "Surgery", chemo ~ "Chemotherapy",
                radiation ~ "Radiation", state~ "State", her2therapy ~ "HER2 treatment",
                grade ~ "Grade", pvgroup ~ "Gene group")
gtsummary::tbl_regression(m_fixed,exponentiate=TRUE,label=my_labels)
```
Note the decreasing hazard ratio with year of diagnosis that is a product of the spread of genetic testing to lower clinical risk groups as discussed as part of the variable selection rationale in @model-strat above.
:::

##### Tests of heterogeneity

```{r}
fvar<-format(as.list(frailtyEM:::summary.emfrail(m_emfrail)$fr_var),scientific=FALSE)
ca<-format(as.list(frailtyEM:::summary.emfrail(m_emfrail)$ca_test),scientific=FALSE)
lrt<-format(as.list(frailtyEM:::summary.emfrail(m_emfrail)$loglik),scientific=FALSE)

p11<-table_fig(fvar,tbl_title="Variance",.9)
p12<-table_fig(ca,tbl_title="Commenges-Andersen test",1)
p13<-table_fig(lrt,tbl_title="Likelihood ratio test",.8)
cowplot::plot_grid(p11,p12,p13,nrow=1)
```

### Breast *ER/PR+, HER2-*

::: {.callout-note collapse="true"}
#### Sample description and missing data (click to view)

```{r}
covar3<-covar1
can_type=3  # use this code for cancer_types 2,3,5
covar<-eval(as.name(paste0("covar",can_type,sep="")))
mort<-get_data(cancer_type=can_type)
# mylabels as defined above for triple negative dz
gtsummary::tbl_summary(mort,label=my_labels, missing_text="missing",include=covar) |>
    gtsummary::modify_caption("Table 1. Covariates - ER/PR+,HER2-") |>
    gtsummary::modify_footnote(label = "Omitted category listed first for categorical variables" )
```
:::

#### Model output

##### Random effects for gene group (as shown in figure 1)

```{r}
#| message: false
# random effects
if(params$frailty=="on") {
  m_emfrail<-get_random(mort,can_type,covar)
}

# fixed effects
m_fixed<-get_fixed(mort,covar)

# graph
## process fixed effects
fe<-process_fixed(m_fixed)
## process random effects
file_name=paste("data/","m_emfrail",can_type,".Rdata",sep="")
load(file_name)
logz<-process_random(m_emfrail)
## combine and graph
datt<-combine_effects(fe,logz)
#plot(graph_effects_vert(datt)) 
p3<-graph_effects_vert(datt)
p3 <- forestploter::edit_plot(p3,
               col = 1,
               row = c(1:3,7),
               gp = gpar(fontface = "italic"))
p3 <- forestploter::insert_text(p3, text = expression(bold(paste("ER/PR-Positive",bolditalic(" HER2"),"-Negative Breast Cancer"))),
              part = "header",
              col = 1:3)
my_pval<-1/2*pchisq(frailtyEM:::summary.emfrail(m_emfrail)$loglik[[3]],1, lower.tail=FALSE)
my_chi<-paste0("Likelihood ratio test for heterogeneity \U{03A7}\U{00B2} = ",round(frailtyEM:::summary.emfrail(m_emfrail)$loglik[[3]],2),", p",pvalr(my_pval))
p3 <- forestploter::insert_text(p3, text = my_chi,
   part = "header", just = "right",
   col = 3, gp = gpar(fontsize=10) )

ggsave(p3,filename=paste("img/","m_emfrail",can_type,".","svg",sep=""))
ggsave(p3,filename=paste("img/","m_emfrail",can_type,".","pdf",sep=""),device=cairo_pdf)
```

![](img/m_emfrail3.svg)

::: {.callout-note collapse="true"}
##### Fixed effects model (click to view)

```{r}
my_labels<-list(agedec ~ "Age (in decades)", race ~ "Race", poverty ~ "Poverty",
                rurality ~ "Rurality", married ~ "Married", yeardx_cat ~ "Year of diagnosis",
                stage ~ "Stage", surgcat ~ "Surgery", chemo ~ "Chemotherapy",
                radiation ~ "Radiation", state~ "State", her2therapy ~ "HER2 treatment",
                grade ~ "Grade", pvgroup ~ "Gene group")
gtsummary::tbl_regression(m_fixed,exponentiate=TRUE,label=my_labels)
```

Note the decreasing hazard ratio with year of diagnosis that is a product of the spread of genetic testing to lower clinical risk groups as discussed as part of the variable selection rationale in @model-strat above.
:::

##### Tests of heterogeneity

```{r}
fvar<-format(as.list(frailtyEM:::summary.emfrail(m_emfrail)$fr_var),scientific=FALSE)
ca<-format(as.list(frailtyEM:::summary.emfrail(m_emfrail)$ca_test),scientific=FALSE)
lrt<-format(as.list(frailtyEM:::summary.emfrail(m_emfrail)$loglik),scientific=FALSE)

p11<-table_fig(fvar,tbl_title="Variance",.9)
p12<-table_fig(ca,tbl_title="Commenges-Andersen test",1)
p13<-table_fig(lrt,tbl_title="Likelihood ratio test",.8)
cowplot::plot_grid(p11,p12,p13,nrow=1)
```

## Colorectal

::: {.callout-note collapse="true"}
### Sample description and missing data (click to view)

```{r}
covar4<-c("agedec","race","sex","poverty","rurality","married","yeardx_cat","stage",
          "surgcat","chemo","radiation","state","colon_rectal","grade","pvgroup")  #colorectal
can_type=4   
covar<-eval(as.name(paste0("covar",can_type,sep="")))
mort<-get_data(cancer_type=can_type) 

my_labels<-list(agedec ~ "Age (in decades from mean age)", race ~ "Race", sex ~ "Sex",
                poverty ~ "Poverty", rurality ~ "Rurality", married ~ "Married", 
                yeardx_cat ~ "Year of diagnosis", stage ~ "Stage", 
                surgcat ~ "Surgery (Yes=1)", chemo ~ "Chemotherapy (Yes=1)", 
                radiation ~ "Radiation (Yes=1)", state~ "State", 
                colon_rectal ~ "Colorectal (Rectal=1)", grade ~ "Grade", 
                pvgroup ~ "Gene group")
gtsummary::tbl_summary(mort,label=my_labels, missing_text="missing",include=covar) |>
    gtsummary::modify_caption("Table 1. Covariates - Colorectal")  |>
    gtsummary::modify_footnote(label = "Omitted category listed first for categorical variables" )

```
:::

### Model output

#### Random effects for gene group (as shown in figure 1)

```{r}
#| message: false

# random effects
if(params$frailty=="on") {
  m_emfrail<-get_random(mort,can_type,covar)
}

# fixed effects
m_fixed<-get_fixed(mort,covar)

# graph
## process fixed effects
fe<-process_fixed(m_fixed)
## process random effects
file_name=paste("data/","m_emfrail",can_type,".Rdata",sep="")
load(file_name)
logz<-process_random(m_emfrail)
## combine and graph
datt<-combine_effects(fe,logz)
p4<-graph_effects_vert(datt)
p4 <- forestploter::edit_plot(p4,
               col = 1,
               row = c(1:3,7),
               gp = gpar(fontface = "italic"))
p4 <- forestploter::insert_text(p4, text = "Colorectal Cancer",
              part = "header",
              col = 1:3,
              gp = gpar(fontface = "bold"))
my_pval<-1/2*pchisq(frailtyEM:::summary.emfrail(m_emfrail)$loglik[[3]],1, lower.tail=FALSE)
my_chi<-paste0("Likelihood ratio test for heterogeneity \U{03A7}\U{00B2} = ",round(frailtyEM:::summary.emfrail(m_emfrail)$loglik[[3]],2),", p",pvalr(my_pval))
p4 <- forestploter::insert_text(p4, text = my_chi,
   part = "header", just = "right",
   col = 3, gp = gpar(fontsize=10) )

ggsave(p4,filename=paste("img/","m_emfrail",can_type,".","svg",sep=""))
ggsave(p4,filename=paste("img/","m_emfrail",can_type,".","pdf",sep=""),device=cairo_pdf)
```

![](img/m_emfrail4.svg)

::: {.callout-note collapse="true"}
#### Fixed effects model (click to view)

```{r}
my_labels<-list(agedec ~ "Age (in decades)", race ~ "Race",   poverty ~ "Poverty",
                rurality ~ "Rurality", married ~ "Married", yeardx_cat ~ "Year of diagnosis",
                stage ~ "Stage", surgcat ~ "Surgery", chemo ~ "Chemotherapy",
                radiation ~ "Radiation", state~ "State", 
                grade ~ "Grade", pvgroup ~ "Gene group")
gtsummary::tbl_regression(m_fixed,exponentiate=TRUE,label=my_labels)
```

Note the decreasing hazard ratio with year of diagnosis that is a product of the spread of genetic testing to lower clinical risk groups as discussed as part of the variable selection rationale in @model-strat above.
:::

#### Tests of heterogeneity

```{r}
fvar<-format(as.list(frailtyEM:::summary.emfrail(m_emfrail)$fr_var),scientific=FALSE)
ca<-format(as.list(frailtyEM:::summary.emfrail(m_emfrail)$ca_test),scientific=FALSE)
lrt<-format(as.list(frailtyEM:::summary.emfrail(m_emfrail)$loglik),scientific=FALSE)

p11<-table_fig(fvar,tbl_title="Variance",.9)
p12<-table_fig(ca,tbl_title="Commenges-Andersen test",1)
p13<-table_fig(lrt,tbl_title="Likelihood ratio test",.8)
cowplot::plot_grid(p11,p12,p13,nrow=1)
```

## Pancreas

::: {.callout-note collapse="true"}
### Sample description and missing data (click to view)

```{r}
covar5<-c("agedec","race","sex","poverty","rurality","married","yeardx_cat","stage","surgcat","chemo","radiation","state","pvgroup")  #pancreas
can_type=5  # use this code for cancer_types 2,3,5
covar<-eval(as.name(paste0("covar",can_type,sep="")))
mort<-get_data(cancer_type=can_type)

my_labels<-list(agedec ~ "Age (in decades from mean age)", race ~ "Race", sex ~ "Sex",
                poverty ~ "Poverty", rurality ~ "Rurality", married ~ "Married", 
                yeardx_cat ~ "Year of diagnosis", stage ~ "Stage", 
                surgcat ~ "Surgery (Yes=1)", chemo ~ "Chemotherapy (Yes=1)", 
                radiation ~ "Radiation (Yes=1)", state~ "State", 
                pvgroup ~ "Gene group")
gtsummary::tbl_summary(mort,label=my_labels, missing_text="missing",include=covar) |>
    gtsummary::modify_caption("Table 1. Covariates - Pancreas")  |>
    gtsummary::modify_footnote(label = "Omitted category listed first for categorical variables" )
```
:::

For pancreatic cancer the models stage 0 and 1 were collapsed given small sample sizes in those with Stage 0.

### Model output

#### Random effects for gene group (as shown in figure 1)

```{r}
#| message: false
mort$stage<-forcats::fct_collapse(mort$stage, "0-1" = c("0","1"))
# random effects
if(params$frailty=="on") {
    m_emfrail<-get_random(mort,can_type,covar)
}

# fixed effects
m_fixed<-get_fixed(mort,covar)

# graph
## process fixed effects
fe<-process_fixed(m_fixed)
## process random effects
file_name=paste("data/","m_emfrail",can_type,".Rdata",sep="")
load(file_name)
logz<-process_random(m_emfrail)
## combine and graph
datt<-combine_effects(fe,logz)
p5<-graph_effects_vert(datt)
p5 <- forestploter::edit_plot(p5,
               col = 1,
               row = c(1:3,7),
               gp = gpar(fontface = "italic"))
p5 <- forestploter::insert_text(p5, text = "Pancreatic Cancer",
              part = "header",
              col = 1:3,
              gp = gpar(fontface = "bold"))
my_pval<-1/2*pchisq(frailtyEM:::summary.emfrail(m_emfrail)$loglik[[3]],1, lower.tail=FALSE)
my_chi<-paste0("Likelihood ratio test for heterogeneity \U{03A7}\U{00B2} = ",round(frailtyEM:::summary.emfrail(m_emfrail)$loglik[[3]],2),", p",pvalr(my_pval))
p5 <- forestploter::insert_text(p5, text = my_chi,
   part = "header", just = "right",
   col = 3, gp = gpar(fontsize=10) )

ggsave(p5,filename=paste("img/","m_emfrail",can_type,".","svg",sep=""))
ggsave(p5,filename=paste("img/","m_emfrail",can_type,".","pdf",sep=""),device=cairo_pdf)
```

![](img/m_emfrail5.svg)

::: {.callout-note collapse="true"}
#### Fixed effects model (click to view)

```{r}
my_labels<-list(agedec ~ "Age (in decades)", race ~ "Race",   poverty ~ "Poverty",
                rurality ~ "Rurality", married ~ "Married", yeardx_cat ~ "Year of diagnosis",
                stage ~ "Stage", surgcat ~ "Surgery", chemo ~ "Chemotherapy",
                radiation ~ "Radiation", state~ "State", 
                pvgroup ~ "Gene group")
gtsummary::tbl_regression(m_fixed,exponentiate=TRUE,label=my_labels)
```

Note the decreasing hazard ratio with year of diagnosis that is a product of the spread of genetic testing to lower clinical risk groups as discussed as part of the variable selection rationale in @model-strat above.
:::

#### Tests of heterogeneity

```{r}
fvar<-format(as.list(frailtyEM:::summary.emfrail(m_emfrail)$fr_var),scientific=FALSE)
ca<-format(as.list(frailtyEM:::summary.emfrail(m_emfrail)$ca_test),scientific=FALSE)
lrt<-format(as.list(frailtyEM:::summary.emfrail(m_emfrail)$loglik),scientific=FALSE)

p11<-table_fig(fvar,tbl_title="Variance",.9)
p12<-table_fig(ca,tbl_title="Commenges-Andersen test",1)
p13<-table_fig(lrt,tbl_title="Likelihood ratio test",.8)
cowplot::plot_grid(p11,p12,p13,nrow=1)
```

## Arrange plots

Extract legend from dummy plot

```{r}
p_leg<-ggplot(
  datt, aes(x = ord, y = estimate, group = as.factor(type), 
            color = as.factor(type)) ) + 
  geom_point() + geom_line()  +
  scale_color_manual(values = c("#377eb8", "#4daf4a"), name="Estimates", 
                     labels=c("Fixed", "Random"),
                     guide = guide_legend(override.aes = list(size = 3,shape=22,
                                                              fill=c("#377eb8", "#4daf4a")) )
                     )  +
  theme(legend.key = element_rect(fill = "white"), 
        legend.text=element_text(size=12),legend.title=element_text(size=12))

extract_legend<-function(myggplot){
    tmp <- ggplot_gtable(ggplot_build(myggplot))
    leg <- which(sapply(tmp$grobs, function(x) x$name) == "guide-box")
    legend <- tmp$grobs[[leg]]
    return(legend)
}
my_legend<-extract_legend(p_leg)
#cowplot::plot_grid(p1,my_legend)
```

Plot 5 figures and legend together for Figure 1.

```{r}
p_final<-cowplot::plot_grid(p1,p2,p3,p4,p5,my_legend, ncol=2,
                            align="AUTO")
save(p_final,file="data/p_final.Rdata")
ggplot2::ggsave(p_final, 
                filename = "final_plot.pdf", 
                path = "img/", 
                units = "in", width =11.6, height = 15, 
                dpi = 300, bg = "white",device=cairo_pdf)
```

```{r}
#| echo: false
knitr::knit_exit()
```

## Extra

```{r}
load("data/p_final.Rdata")
ggplot2::ggsave(p_final, 
                filename = "final_plot.pdf", 
                path = "img/", 
                units = "in", width =11.6, height = 15, 
                dpi = 300, bg = "white",device=cairo_pdf)
```

## Text

Given that we were estimating individual effects for 7-8 genetic exposures over 5 cancers there were up to 40 estimates of primary interest. To reduce the effect of small samples and account for multiple comparisons we used frailty models to estimate the gene comparisons in each cancer model. We used a gamma frailty model. We followed Balan(2020) in estimating the random and fixed effects for the gene effects. As in Balan, both the fixed and random effects are centered around 0 for comparability.

1.  Balan TA, Putter H. A tutorial on frailty models. Stat Methods Med Res. 2020 Nov;29(11):3424–3454. PMCID: PMC7534210

Figure X. Hazard ratios for commonly found genes in 5 cancers Figure Legend: The figure shows hazard ratios for cancer-specific mortality for 7 or 8 groups of people representing a set of more commonly found pathogenic variants (*ATM, CHEK2, PALB2, BRCA 1/2*, Lynch syndrome genes(5)) and groups representing those with other PVs, variants of unknown significance(VUS) or no pathogenic variants. Separate models were run for each cancer. Given the multiple comparisons and heterogeneity in the number of people with different PVs estimates are presented for models with the gene group estimates as fixed effects (dummy variables) and as empirical Bayes estimates from frailty models. We used a gamma frailty model and followed Balan(2020) in estimating the random and fixed coefficients. As in Balan, both the fixed and random coefficients are centered around 0 and thus the hazard ratios shown represent a comparison to a hypothetical person with the average hazard across all of the genetic groups in this population holding other covariates in the model constant.

Tried this for debugging runs in chunks but not in knit

```{r}
library(data.table)
can_type=1
file_name=paste("data/","m_emfrail",can_type,".Rdata",sep="")
load(file_name)
data<-summary(m_emfrail)[[12]]
data
#arg<-summary(m_emfrail)[[12]]
#arg
#process_random3(arg)
```

Other graphs

```{r}
surv_1<-survminer::ggforest(m_fixed, data = as.data.frame(mort),
                    fontsize = .8)
ggsave(plot=surv_1,
      filename = "surv_1.pdf", 
      path = "img/", 
      units = "in", width =11, height = 14, dpi = 300, bg = "white")
```

```{r}
test<-broom::tidy(m_fixed,conf.int=TRUE,exponentiate=TRUE)
forester::forester(left_side_data=test[,1],
                   estimate=test$estimate,
                   ci_low=test$conf.low,
                   ci_high=test$conf.high,
                   xlim=c(-4,4))
```

```{r}

ggplot(broom::tidy(survival::survfit(m_fixed)), aes(time, estimate)) +
  geom_line() +
  geom_ribbon(aes(ymin = conf.low, ymax = conf.high), alpha = .25)

```

Less ideal

``` r
library(gridExtra)
p_final<-grid.arrange(p,p,p,p,p,p5,ncol=2)
```

``` r
# doesn't work
library(patchwork)
p1 + p2
```
